Linear vertical shear instability in astrophysical disks: a numerical study

### 2014-11-28 linear_iso.tex
minor update

\subsection{Neccessity of vertical shear for instability}
Writing
\begin{align}
  V \equiv Q\exp{\left[-\frac{\hat{z}^2}{2}(\ii q h
      \hat{k})\right]},
\end{align}
we have an alternative form of the governing equation 
\begin{align}\label{iso_alt}
  \frac{d}{d\hat{z}}\left[w(\hat{z})\left(\frac{dV}{d\hat{z}} + \ii q h
    \hat{k}\hat{z}V\right)\right] + n^2V w(\hat{z}) =0, 
\end{align}
where 
\begin{align}
  w(\hat{z}) \equiv \exp{\left(-\frac{\hat{z}^2}{2}\right)},  
\end{align}
and we have defined a new eigenvalue
\begin{align}
  n^2 \equiv \left(\frac{\sigma}{\Omega_k}\right)^2(1+\hat{k}^2). 
\end{align}

Next, we multiply Eq. \ref{iso_alt} by $V^{*}$, integrate
vertically and assume boundary terms vanish to obtain
\begin{align}\label{integral_relation}
  n^2\int_{-\infty}^{\infty}w|V|^2d\hat{z} 
  =&\int_{-\infty}^{\infty}w\left|\frac{dV}{d\hat{z}}\right|^2d\hat{z}\notag\\
  +&\ii q h \hat{k}\int_{-\infty}^{\infty}w\hat{z}V\frac{dV^*}{d\hat{z}}d\hat{z}.   
\end{align}
It follows that for instability ($\imag\sigma>0$ and therefore $\imag
n^2 \neq 0$), it is neccessary to have $q\neq 0$, i.e. vertical
shear. We will use this integral relation to estimate growth
rates.

\subsection{Stability in the absence of vertical shear}
When $q=0$, the governing equation is
\begin{align}
  \frac{d}{d\hat{z}}\left[w(\hat{z})\frac{dV}{d\hat{z}}\right] + n^2V
  w(\hat{z}) =0, 
\end{align}
i.e. Hermite's differential equation. For bounded solutions we have 
\begin{align}
  V\propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to
\begin{align}
  \left|\frac{\sigma}{\Omega_k}\right| = n
  \left(1+\hat{k}^2\right)^{-1/2}. 
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2$, our analysis is only
valid for large wavenumbers such that $\hat{k}^2\gg 
n^2-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height. From

@@

Linearizing the integral Eq. \ref{integral_relation} and taking the
imaginary part, we have

>>>>

\subsection{Neccessity of vertical shear for instability}
Writing
\begin{align}
  V \equiv Q\exp{\left[-\frac{\hat{z}^2}{2}(\ii q h
      \hat{k})\right]},
\end{align}
we have an alternative form of the governing equation 
\begin{align}\label{iso_alt}
  \frac{d}{d\hat{z}}\left[w(\hat{z})\left(\frac{dV}{d\hat{z}} + \ii q h
    \hat{k}\hat{z}V\right)\right] + nV w(\hat{z}) =0, 
\end{align}
where 
\begin{align}
  w(\hat{z}) \equiv \exp{\left(-\frac{\hat{z}^2}{2}\right)},  
\end{align}
and we have defined a new eigenvalue
\begin{align}
  n \equiv \left(\frac{\sigma}{\Omega_k}\right)^2(1+\hat{k}^2). 
\end{align}

Next, we multiply Eq. \ref{iso_alt} by $V^{*}$, integrate
vertically and assume boundary terms vanish to obtain
\begin{align}\label{integral_relation}
  n \int_{-\infty}^{\infty}w|V|^2d\hat{z} 
  =&\int_{-\infty}^{\infty}w\left|\frac{dV}{d\hat{z}}\right|^2d\hat{z}\notag\\
  +&\ii q h \hat{k}\int_{-\infty}^{\infty}w\hat{z}V\frac{dV^*}{d\hat{z}}d\hat{z}.   
\end{align}
It follows that for instability ($\imag\sigma>0$ and therefore $\imag
n \neq 0$), it is neccessary to have $q\neq 0$, i.e. vertical
shear. We will use this integral relation to estimate growth
rates.

We also recognize that in Eq. \ref{integral_relation}, 
the factor $q\hat{z}$ represents vertical shear and $w(\hat{z})$ is
the background density profile. Assuming $V$ is either odd or even in
$\hat{z}$,  then for the second integral on the right
hand side to be non-zero, it is neccessary to have a vertical shear
profile that has odd symmetry in the vertical co-ordinate.

\subsection{Stability in the absence of vertical shear}
When $q=0$, the governing equation is
\begin{align}
  \frac{d}{d\hat{z}}\left[w(\hat{z})\frac{dV}{d\hat{z}}\right] + nV
  w(\hat{z}) =0, 
\end{align}
i.e. Hermite's differential equation. For bounded solutions we have 
\begin{align}
  V\propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to
\begin{align}
  \left|\frac{\sigma}{\Omega_k}\right| = \sqrt{n}
  \left(1+\hat{k}^2\right)^{-1/2}. 
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2$, our analysis is only
valid for large wavenumbers such that $\hat{k}^2\gg 
n-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height. From

@@

Linearizing the integral Eq. \ref{integral_relation} and taking the
imaginary part, we have
\begin{align}
  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
  \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
  \He_n^2(\hat{z}) d\hat{z} \notag\\
  = \delta q h \hat{k} 
  \int_{-\infty}^{\infty}
  w(\zhat)\hat{z}\He_n(\hat{z})\frac{d\He_n}{d\zhat} d\hat{z}
\end{align}
Integrating the right hand side by parts and recognizing $\He_2 =
\hat{z}^2 -1$, we have
\begin{align}
  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
  \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
  \He_n^2(\hat{z}) d\hat{z} \notag\\
  = \frac{1}{2}\delta q h \hat{k} 
  \int_{-\infty}^{\infty}
  w(\zhat)\He_2(\hat{z})\He_n^2(\hat{z})d\hat{z}. 
\end{align}
Finally, we evaluate the integrals using following properties of Hermite polynomials
\begin{align}
  \int_{-\infty}^{\infty}
  w(\xi) \He_k(\xi)\He_l(\xi) d\xi = \sqrt{2\pi}k!\delta_{kl}, 
\end{align}
and
\begin{align}
  \int_{-\infty}^{\infty}
  w(\xi) \He_m(\xi)\He_k(\xi)\He_l(\xi) d\xi =
  \frac{\sqrt{2\pi}m!k!l!}{(s-m)!(s-k)!(s-l)!}, 
\end{align}
where $s = (m+k+l)/2$. We then obtain
\begin{align}
 \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
 \left(1+\hat{k}^2\right) = \delta q h \hat{k} \frac{n!}{(n-1)!}.
\end{align}
Inserting the original eigenfrequency $\sigma = \pm
\sqrt{n}\Omega_k(1+\hat{k}^2)^{-1/2}$ gives
\begin{align}
  \frac{\imag(\delta\sigma)}{\Omega_k} = \pm \frac{1}{2}\delta q h
  \sqrt{n} \frac{\hat{k}}{\sqrt{1+\hat{k}^2}}. 
\end{align}
 For $\hat{k}\gg 1$ we have
\begin{align}
  \frac{\imag(\delta\sigma)}{\Omega_k} \simeq \pm \frac{1}{2}\delta q h
  \sqrt{n} \sgn{\hat{k}}. 
\end{align}
